Topological Equivariant Artist Model for Visualization Library Architecture

### 2023-05-08 paper/main.tex
main
intro take 1000

\note{The mapping between data and visualization is expected to be bidirectional -> that's what we really mean by equivariance and homomorphism  }
\IEEEPARstart{V}isualizations are expected to be structure preserving, meaning that when data is transformed into a graphical representation enough of the structure of the data is present in the graphic that the data can be extracted from the graphical representation. Changes to structure through changing the data or the graphic in the $data\leftrightarrow graphic$ mapping are expected to be reflected in each inputs corresponding output \cite{kindlmannAlgebraicProcessVisualization2014}.Because visualizations are assumed to preserve structure, visual algorithms implemented in visualization producing tools, such as software libraries, are also expected to preserve structure; frequently these expectations are either codified implicitly, only for some aspects of the data structure, or inconsistently with respect to visualization algorithm type. This means that currently many visualization tools lack an explict, comprehensive, and consistent way to express what conditions must be met to gaurantee that the component produced a structure preserving visualization.

We propose that we can encode the structure of the data and graphics in a uniform generalizable manner by using a mathematical structure from algebraic topology called a fiber bundle, and that we can use category theory to explicitly specify what it means for a visualization tool component to preserve structure. These algebraic structures can be translated into analogous programmatic types for encoding structure, while the use of category theory facilitates developing a functional design framework that describes function composition in terms of functional algebraic operations. The typing system and composition inherent in a functional design encourage library developers to build complex visualization components from simple verifiable parts that have few side effects\cite{huHowFunctionalProgramming2015, hughesWhyFunctionalProgramming1989}. These categorically specified components could be built as a standalone library and integrated into existing libraries because they are inherently self-contained. Furthermore we see the application of these ideas in low level visualization library design as our motivation, and we hope these ideas will help shape the reorganization of critical data visualization libraries, such as Matplotlib. The contribution of this paper is a mathematical framework for describing structure preserving visualization components in a manner that is generalizable to different types of inherent structure. This framework also provides guidance for the construction and testing of structure preserving visualization library components.

@@

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|r | l  l |}
\hline
& function & constraint\\
\hline
 \gbasepointc\ to \dbasepointc& $ \vindexc: \opensetgc \rightarrow \opensetc$ & for $\dbasepointc \in \opensetc$ exists $\gbasepointc \in \opensetgc$ \\
 & & s.t. $\vindexc(\gbasepointc) = \dbasepointc$\\
 graphic for \dbasepointc & $ \vindexpushc \gsectionc: \opensetc \rightarrow \vindexpushc\gtotalc\restriction_{\opensetc}$ & $\vindexpushc\gsectionc(\dbasepointc)(\gbasepointc) = \gsectionc(\gbasepointc)$ \\
 record for \gbasepointc & $\vindexpullc\dsectionc: \opensetgc \rightarrow  \vindexpullc\dtotalc\restriction_{\opensetgc}$ & $\vindexpullc\dsectionc(\gbasepointc)=\dsectionc(\vindexc(\gbasepointc)) = \dsectionc(\dbasepointc)$  \\
\hline
  \end{tabular}
  \caption{Functors between graphic and data indexing spaces  \autoref{sec:atct:xi}}
  \label{tab:appendix:summary:transport}
  }
\end{table}

@@

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|r|l|l|}
  \hline
      & function & constraints \\
  \hline
  \textcolor{artist}{artist} & $\vartistc: \cgamma{\dbasec}{\dtotalc} \rightarrow \imartist{\gbasec}{\gtotalc}$ &  \\
  Data to Graphi  &  $\imartist{\gbasec}{\gtotalc} \subset \cgamma{\gbasec}{\gtotalc}$ & $\vindexc(\gbasec) = \dbasec$\\
  & & \\
  \hline 
  Encoders  & & \\
           & & \\
  \hline
  Decompositor & & \\
              && \\
  \hline
\end{tabular}
\caption{artist, verification functions, and construction $\vartist = \vmarkc \circ \nu$ introduced in \autoref{sec:artist}, and \autoref{sec:construction}}
\label{tab:appedendix:summary:artist}
}
\end{table}

>>>>

\IEEEPARstart{V}isualization is a structure preserving mapping from data to graphic. One way of verifying that the structure of the data is present in the graphic is to verify that the data set can be distilled from the graph, i.e. that the values read off a bar chart are equivalent to the values used to construct the bar chart. Since structure preservation requires that the $data\leftrightarrow graphic$ be bidirectional, changes to either data or graphic are expected to propogate to the other side of the transformation\cite{kindlmannAlgebraicProcessVisualization2014}. In this paper, we propose a methodology for describing structure, verifying structure preservation, and specifying the conditions for constructing a structure preserving map between data and graphics. We use category theory, which is the math of structures and their relations, to explicitly specify what it means for a visualization mapping to preserve structure in an explicit, comprehensive, and consistent way. We encode the structure of the data and graphics in a uniform generalizable manner by using a mathematical structure from algebraic topology called a fiber bundle; doing so allows us to discuss everything from lines to complex ND spaces with the same interface.

We use category theory and algebraic structures because they can be directly translated into software design specifications and our goal is to improve visulization tools. The algebraic structures can be translated into analogous programmatic types, while category theory facilitates developing a functional design framework that describes function composition in terms of functional algebraic operations. The typing system and composition inherent in a functional design encourage library developers to build complex visualization components from simple verifiable parts that have few side effects\cite{huHowFunctionalProgramming2015, hughesWhyFunctionalProgramming1989}. These categorically specified components could be built as a standalone library and integrated into existing libraries because they are inherently self-contained. Furthermore we see the application of these ideas in low level visualization library design as our motivation, and we hope these ideas will help shape the reorganization of critical data visualization libraries, such as Matplotlib. The contribution of this paper is a mathematical framework for describing structure preserving visualization components in a manner that is generalizable to different types of inherent structure. This framework also provides guidance for the construction and testing of structure preserving visualization library components.

@@

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.5}
  \begin{tabular}{|r | l  l |}
\hline
& function & constraint\\
\hline
 \gbasepointc\ to \dbasepointc& $ \vindexc: \opensetgc \rightarrow \opensetc$ & for $\gbasepointc \in \opensetgc$ exists $\dbasepointc \in \opensetc$ \\
 & & s.t. $\vindexc(\gbasepointc) = \dbasepointc$\\
 graphic for \dbasepointc & $ \vindexpushc \gsectionc: \opensetc \rightarrow \vindexpushc\gtotalc\restriction_{\opensetc}$ & $\vindexpushc\gsectionc(\dbasepointc)(\gbasepointc) = \gsectionc(\gbasepointc)$ \\
 record for \gbasepointc & $\vindexpullc\dsectionc: \opensetgc \rightarrow  \vindexpullc\dtotalc\restriction_{\opensetgc}$ & $\vindexpullc\dsectionc(\gbasepointc)=\dsectionc(\vindexc(\gbasepointc)) = \dsectionc(\dbasepointc)$  \\
\hline
  \end{tabular}
  \caption{Functors between graphic and data indexing spaces  \autoref{sec:atct:xi}}
  \label{tab:appendix:summary:transport}
  }
\end{table}

@@

\begin{table}[h!]
  \centering
  {\renewcommand{\arraystretch}{1.2}
\begin{tabular}{|r|l|l|}
  \hline
      & function & constraints \\
  \hline
  \textcolor{artist}{artist} & $\vartistc: \cgamma{\dbasec}{\dtotalc} \rightarrow \imartist{\gbasec}{\gtotalc}$ &  \\
  Data to Graphic  &  $\imartist{\gbasec}{\gtotalc} \subset \cgamma{\gbasec}{\gtotalc}$ & $\vindexc(\gbasec) = \dbasec$\\
  & & \\
  \hline 
  Encoders  & & \\
           & & \\
  \hline
  Decompositor & & \\
              && \\
  \hline
\end{tabular}
\caption{artist, verification functions, and construction $\vartist = \vmarkc \circ \nu$ introduced in \autoref{sec:artist}, and \autoref{sec:construction}}
\label{tab:appedendix:summary:artist}
}
\end{table}
